Motion planning for assessing accessibility of protein tunnels

### 2017-09-09 paper.tex
link to web page

%\documentclass[a4paper, 10pt]{IEEEtran}      % for icra/RAL final
\documentclass[usletter, 10pt, conference]{ieeeconf} % initial submission to RAL/ICRA
\usepackage{graphicx}
\usepackage{booktabs}
%\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{subfig}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{multirow}
\usepackage{flushend}
%\usepackage[bookmarks=false]{hyperref}
%\usepackage{setspace}
\pdfminorversion=4

@@

One of the problems in protein engineering is the tunnel detection, where the task is to find a tunnel (pathway) 
leading to a specific place (called active site) inside the protein~\cite{gora2013gates}.
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
%The knowledge about the existence of the tunnels and their properties (e.g., their bottleneck and chemical properties around the tunnel),
%is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
As the interaction takes place in the active site, the existence of a tunnel connection the active site with the outer environment
is necessary for the interaction.
%To get the ligand to the active site, a transportation tunnel connecting the protein outer environment with the active site has to exists.
Such a  tunnel has to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
Chemists are using the tunnel detection method in order to estimate possibility of interactions between a protein and a ligand without need to perform long Molecular Dynamics (MD) simulations or even real experiments (Fig.~\ref{fig::motiv}).
The tunnels are detected mainly using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
a spherical ligand (probe) and they are represented as a sequence of spheres.
However, as the ligands typically non-spherical and flexible,  it is difficult to estimate their traversability through the tunnels computed for only one sphere.
%Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
%This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013}. % or its activity of the protein towards other molecules~\cite{Pavlova2009}.
%The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
%This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
%ligand from the outside environment to the active site or vice versa. 
%Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
%Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
%The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.

@@

In this paper, we propose a modification to RRT to cope with these issues.
In the proposed method, the random samples are not generated from the whole configuration space, but the sampling is guided using
the spheres of the tunnel.
This allows us to find trajectories along the given tunnel, but it also decreases the volume of the configuration space to be searched.
To increase movements of the ligands in the narrow tunnels, the free-space is dilated by 
shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}.
We model the ligand flexibility using a fixed set of predefined conformations.

@@

The proposed approach was used to analyze tunnel traversability towards the active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
The molecular dynamics of the protein was computed using the Amber tool.
%100 frames of the dynamics were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
%The second tunnel is more dynamic, i.e., it fluctuatest more

Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
For each scale, 20 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel (outside the protein).
For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed.

Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), 
    denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners is based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the higher success rates.

@@

The results are shown in Tab.~\ref{tab::one} which shows traversability rate. 
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with bottleneck larger than the tested radius.
The row Planners shows traversable rate detected by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increases the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromised between the performance and the runtime.
The rates provided by \RA\ and \RB\ are similar in this case.

\begin{table}
\centering
\caption{\label{tab::one}Traversability for one atom.}
\small
\renewcommand{\tabcolsep}{3.5pt}
{\small
\input graphProbe.tex
}
\end{table}

@@

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
name shows the number of selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
For the ligand DCP, higher traversability rates was achieved when 50 conformations were selected.
This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using less number of conformations or when preferring the packed conformations.

@@

The most time consuming part of the planners is the collision detection. 
Although both methods can utilize same amount of collision detection queries in the expansion step, \RB\ is slower than \RA.
The reason is, that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
The better performance of \RA\ is caused by the proposed atom-based metric, because this distance between two configurations depends on the actual shape of the ligand.
This shape is different in each configuration.
Contrary, the \RB\ utilizes only the single 6D Euclidean metric which is same for all conformations.
To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
prevent to use KD-tree for the nearest-neighbor search.

>>>>

%\documentclass[a4paper, 10pt]{IEEEtran}      % for icra/RAL final
\documentclass[usletter, 10pt, conference]{ieeeconf} % initial submission to RAL/ICRA
\usepackage{graphicx}
\usepackage{booktabs}
%\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{subfig}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{multirow}
\usepackage{flushend}
\usepackage[bookmarks=false]{hyperref}
%\usepackage{setspace}
\pdfminorversion=4

@@

One of the problems in protein engineering is the tunnel detection, where the task is to find a tunnel (pathway) 
leading to a specific place (called active site) inside the protein~\cite{gora2013gates}.
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
%The knowledge about the existence of the tunnels and their properties (e.g., their bottleneck and chemical properties around the tunnel),
%is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
As the interaction takes place in the active site, the existence of a tunnel connecting the active site with the outer environment
is necessary for the interaction.
%To get the ligand to the active site, a transportation tunnel connecting the protein outer environment with the active site has to exists.
Such a  tunnel has to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
Chemists are using the tunnel detection method in order to estimate possibility of interactions between a protein and a ligand without need to perform long Molecular Dynamics (MD) simulations or even real experiments (Fig.~\ref{fig::motiv}).
The tunnels are detected mainly using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
a spherical ligand (probe) and they are represented as a sequence of spheres.
However, as the ligands typically non-spherical and flexible,  it is difficult to estimate their traversability through the tunnels computed for only one sphere.
%Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
%This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013}. % or its activity of the protein towards other molecules~\cite{Pavlova2009}.
%The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
%This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
%ligand from the outside environment to the active site or vice versa. 
%Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
%Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
%The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.

@@

In this paper, we propose a modification to RRT to cope with these issues.
The random samples are not generated from the whole configuration space, but the sampling is guided using
the spheres of the tunnel.
This allows us to find trajectories along the given tunnel, which is faster than searching any trajectory in the whole protein.
To increase movements of the ligands in the narrow tunnels, the free-space is dilated by 
shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}.
The ligand flexibility is modelled using a fixed set of predefined conformations.

@@

The proposed approach was used to analyze tunnel traversability towards the active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
The molecular dynamics of the protein was computed using the Amber tool.
%100 frames of the dynamics were analyzed with the aim to calculate percentage of frames where the active site was accessible through the tunnels.
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.~\footnote{%
We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional images.
}
The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.

Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
For each scale, 20 collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
For each ligand, each minimal scale $\smin$ and each initial configuration, 10 trajectories were computed.

Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), 
    denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction} denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA.
Both methods were implemented in the same software framework and they utilized the same data structures and routines for collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners is based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the higher success rates.

@@

The results are shown in Tab.~\ref{tab::one} which shows traversability rate. 
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with bottleneck larger than the tested radius.
The row Planners shows traversable rate detected by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increases the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
The rates provided by \RA\ and \RB\ are similar in this case.

\begin{table}
\centering
\caption{\label{tab::one}Traversability for one atom.}
\renewcommand{\tabcolsep}{3.5pt}
{\small
\input graphProbe.tex
}
\end{table}

@@

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
name shows the number of selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
For the ligand DCP, a higher traversability rate was achieved when 50 conformations were selected.
This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using less number of conformations or when preferring the packed conformations.

@@

The most time consuming part of the planners is the collision detection. 
Although the both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
The reason is, that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
The better performance of \RA\ is caused by the proposed atom-based metric, because this distance between two configurations depends on the actual shape of the ligand.
This shape is different in each configuration.
Contrary, the \RB\ utilizes only the single 6D Euclidean metric which is same for all conformations.
To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
prevent the usage KD-tree for the nearest-neighbor search.
